---
title: "Sampling the Imaginary"
author: "Florian Sckade"
output: 
  html_document:
   theme: cosmo
   highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Disclaimer: These are paraphrased or even direct quotes from the book. 

Bayesian inference is distinguished by a broad view of probability, not by the use of
Bayes' theorem. A central part explained here is the working with samples from the 
posterior distribution, to make inference in the style of questions for the defined
model easier. This is opposed to the classic integral calculus you utilize when 
usually working with distribution functions. With this method, instead of using these
mathematical formulas for integration, you can just use standard data summary methods-

My favorite quote from the first section of the chapter:  

> "No one has ever seen a probability"

#### **Rethinking:** Hypothesis-Testing

Scientific inference process: (1) An hypothesis is true or false; (2) we calculate a 
statistical measure to determine if the hypothesis is likely to be true/false; (3) one
*should* use Bayes theorem to deduce the impact of our measurement on the original 
hypothesis. 
The example: suppose the probability of a positive finding when the hypothesis is true 
(*power*), is Pr(sig$|$true)=0.95, and that the probability of a positive finding 
when the hypothesis is false is Pr(sig$|$false)=0.05 (*false positive rate*, like the
conventional 5% in significance testing). Additionally, suppose that 1 in every 100 
hypothesis is true, Pr(true)=0.01. 
Thus, in this scenario, we can calculate Pr(true$|$pos)=0.16, meaning a positive finding
corresponds to a probability of 16% that the hypothesis is true. An shrinking of the 
false-positve rate to 1% does not solve this problem, merely improving the posterior
probability to a 0.5. Pr(true) is the most essential rate in this setting, for which
thinking is required, not testing. 

## Sampling to summarize

Once a posterior distribution is estimated, samples can be drawn from it. In the first 
place, this only recreates the density function we already calculated. However, these
samples can also then be used to *describe* and *understand* the posterior, which is very
important. 

Common questions for summaries of posterior samples:

* How much posterior probability lies below some parameter value?
* How much posterior probability lies between two parameter values?
* Which parameter value marks the lower 5% of the posterior probability?
* Which range of parameter values contains 90% of the posterior probability?
* Which parameter value has highest posterior probability?

These questions can be divided into:  

1. questions about intervals of defined boundaries, 
2. questions about intervals of defined probability mass
3. questions about point estimates. 

Essentially, there are different types of intervals you can utilize to analyze the posterior. 
For (1), this means setting an interval based on specific parameter values and analyzing,
how much probability mass lies between these boundaries.
In terms of defines mass (2), one can build *compatibility intervals*. These posterior
intervals report two parameter values that contain between them a specified amount of 
posterior probability. 
Point estimates (3) depend on more information, as to what kind of summary statistic should 
actually be utilized. The bayesian parameter estimate is precisely the posterior distribution,
and as such, the decision on a point estimate to summarize this distribution is not trivial.
When deciding between different point estimates (mode, mean, median), one can use *loss functions*. 
Examples are *absolute loss*(leads to median) and *quadratic loss* (leads to posterior mean).

## Sampling to simulate prediction

Samples can be used to generate implied observations, which is useful because of:

1. *Model Design*: Sample the prior to understand its implications.
2. *Model checking*: Investigate model behavior after updating and check if it worked correctly.
3. *Software validation*: Check if model fitting software works as expected.
4. *Research design*: Simulate observations from the hypothesis, to evaluate research design. 
                      (akin to power analysis)
5. *Forecasting*: Samples can be used as predictions, but also as model criticism. 


# Exercises

```{r}
p_grid <- seq(from = 0, to = 1, length.out = 1000)
prior <- rep(1, 1000)
likelihood <- dbinom(6, size = 9, prob = p_grid)
posterior <- likelihood * prior 
posterior <- posterior / sum(posterior)

set.seed(100)
samples <- sample(p_grid, prob = posterior, size = 1e4, replace = TRUE)
```


#### 3E1. How much posterior probability lies below p = 0.2?
```{r}
# Either this way
sum(samples < 0.2) / 1e4
# or more directly:
mean(samples < 0.2)
```
#### 3E2. How much posterior probability lies above p = 0.8?
```{r}
mean(samples > 0.8)
```
#### 3E3. How much posterior probability lies between p = 0.2 and p = 0.8?
```{r}
mean(samples > 0.2 & samples < 0.8)
```
#### 3E4. 20\% of the posterior probability lies below which value of p?
```{r}
quantile(samples, 0.2)
```
#### 3E5. 20\% of the posterior probability lies above which value of p?
```{r}
quantile(samples, 0.8)
```
#### 3E6. Which values of p contain the narrowest interval equal to 66\% of the posterior probability?
```{r}
# Highest posterior density interval
rethinking::HPDI(samples, prob = 0.66)
```
#### 3E7. Which values of p contain 66\% of the posterior probability, assuming equal posterior probability both below and above the interval? 
```{r}
# Percentile interval
rethinking::PI(samples, prob = 0.66)
```
#### 3M1. 
Suppose the globe tossing data had turned out to be 8 water in 15 tosses. 
Construct the posterior distribuition, using grid approximation. Use the
same flat prior as before. 
```{r}
p_grid <- seq(from = 0, to = 1, length.out = 1000)
prior <- rep(1, 1000)
likelihood <- dbinom(8, size = 15, prob = p_grid)
posterior <- likelihood * prior 
posterior <- posterior / sum(posterior)
plot(x = p_grid, y = posterior, type = "l")
```
#### 3M2
Draw 10,000 samples from the grid approximation from above. Then use the
samples to calculate the 90 \% HDPI for p. 
```{r}
samples <- sample(p_grid, prob = posterior, size = 1e4, replace=TRUE)
rethinking::HPDI(samples, prob = 0.9)
```
#### 3M3 
Construct a posterior predictive check for this model and data.
This means simulate the distribution of samples, averagin over
the posterior unvertanity in p. What is the probability of observing
8 water in 15 tosses? 
```{r}
w <- rbinom(1e4, size = 15, prob = samples)
mean(w == 8)
```

